Diversity and evolution of centromere repeats in the maize genome

### 2014-05-03 centc.tex
sleepy finished work on discussion. will revisit tomorrow with more brain power

\jri{check tense throughout. i think some is past and some is present} 
Our analysis of centromere repeat diversity across the maize genome identifies thousands of copies exhibiting tremendous diversity. 
Nonetheless, while we can cluster the repeats into groups of related sequences, these groups seem to have little relation to current or ancient maize chromosomes (Figure \ref{pcoa}).  
We see no evidence of chromosome specific repeats such as that seen in X and Y \citep{Kawabe2005, Pontes2004}, with the currently available evidence suggesting that the chromosome 4 repeats previously called Cent4 is in fact a poorly characterized retrotransposon.

Instead of chromosome-specific groups, we find that virtually all the large arrays of CentC in the maize  genome derive from the maize1 subgenome, which is known to have lower gene loss and higher average expression than the maize subgenome 2 \cite{Schnable2011}. \jri{is it maize1 or maize subgenome 1? you use both here} 
One possible explanation for this clear bias is that differential expression between subgenomes affects centromere function.  Regulatory microRNA’s corresponding to centromeric repeats have been observed\citep{ReinhartBartel2002}, and perhaps the silenced centromere became more susceptible to deletion just like other parts of subgenome2.  We see no evidence of a more rapid decay of CentC sequences in subgenome2, but we likely have poor ability to detect deletion of entire tandem arrays.  An alternative explanation is simply that the subgenome1 parent had more or larger CentC arrays, leading to an unequal representation in the ancient polyploid ancestor of maize.

@@

%!RELATEDNESS
We also explored how genetic relatedness between CentC’s correlated with location along the physical map.  According to our PCoA analyses, CentC copies across the genome do not form distinct genetic groups correlating with past origin (Figure \ref{pcoa}).  The lack of subgenome or chromosome grouping in modern CentC suggests a little differentiation between the repeats since the ancestral divergence.  Given the lack of a genome wide pattern, we also wanted to investigate genetic relatedness on individual chromosomes.  We hypothesized that most copies of CentC arose from local duplication rather than transposition, and therefore the genetic distance between CentC’s would be lowest across CentC’s within a large tandem array.  \pb{This part is ugly, and needs help} From the reference genome, we chose to investigate chromosomes 2 and 5, since they have been sequenced from end to end, allowing for inferences about the relationship between diversity and physical proximity \citep{Wolfgruber2009}.  Furthermore, the role of the repeat arrays on chromosomes 2 and 5 appear very different, as the largest array on chromosome 2 interacts with the kinetochore, while no array on chromosome 5 does \citep{Wolfgruber2009}.  We find that repeats on chromosomes 2 and 5 are most highly genetically related to neighboring copies (Figure \ref{heatmap}).  This relationship was recapitulated in analyses using SpaGeDi, where CentC’s within approximately 10-50KB of one another were more genetically similar (Tables \ref{Supp_SpagediChr02} and \ref{Supp_SpagediChr05}).  Though we did not investigate the relationship between CentC location and genetic similarity on the other chromosomes due to incomplete sequencing of centromeres, we observe that many CentC’s within an array fall into the same  significant grouping in our Tracy-Widom clustering analysis, suggesting that high local similarity of CentC’s is a genome-wide phenomenon.  This local relatedness suggests that most CentC is evolving through tandem duplication and not a long distance mechanism such as transposition that had been previously suggested \citep{Shi2010}.  Concerning chromosomes 2 and 5, the higher local relatedness of all CentC clusters regardless of presence within the kinetochore suggests that CentC interaction with kinetochore proteins does not have a detectable change on their rates of evolution.

When investigating why a pair of chromosomes on two different chromosomes are each other closest genetic relative, we revealed that homoplasy in mutations is common across CentC variants.  We suggest that copies of CentC are sufficiently old within the genome that homoplasious mutations cause physically distant CentC’s to be highly genetically related.  Importantly, we also do not see PCoA group capturing full clusters across chromosomes in a way that would be consistent with retrotransposition. We speculate that a vast majority of the CentC’s exist as a result of very old duplications meaning that mutations have had a long time to accumulate. Roughly 80\% of the CentC repeats have their closest genetic relative on the same chromosome, an observation we would expect if CentC’s on a chromosome share ancestry.  However, only 14\% of closest genetic pairs are found within 10KB of each other, suggesting that their tandem duplication is old and that most CentC’s have persisted within the genome for a long time. \pb{This is an important sentence, and I can’t seem to get it to communicate my point properly.  perhaps because im making too many points}

%!DOMESTICATION
\jri{insert CITE, italics, probably refs to tables/figs as appropriate}  Previous cytogenetic work  identified  differences in centromere repeat content  between domesticated maize and its wild relatives Z. mays ssp. parviglumis,), Z. mays ssp. mexicana,, and Z. luxurians (hereafter parviglumis, mexicana, and luxurians) \citep{Albert2010} but was unable to quantify these differences. Our resequencing results show that while there is little difference in the distribution of CentC in tandem arrays, the total abundance of CentC has  decreased  \jri{can we say absolute abundance? we have genome size estimates, right?} during domestication, and we verify this with FISH in two maize-teosinte hybrid individuals (Figure \ref{abundance}). This result contrasts with the apparent increase in transposable element abundance  \citep{Chia2012}, suggesting that changes in CentC are not due to causes common to all repetitive sequence. \jri{ are we sure TEs have increase in absolute number? didn't chia et al. look at RPKM?  wouldn't that increase if overall genome size decreases? if we control for genome size, does TE content show and absolute increase?} The maize genome is smaller than its teosinte counterpart, largely due to differences in the abundance of heterochromatic knobs (Poggio et al 1998).  \citet{Zhang2012} have postulated an adaptive relationship between centromere size and genome size based on an observed correlation between centromere size and genome size across a number of grass species.  This model would propose that the decrease in CentC abundance seen post-domestication is due to selection for smaller active centromeres to complement the smaller overall genome size.  While our current data are insufficient to evaluate this conclusion, future work investigating differences in CentC content among maize landraces that vary in genome size (CITE) may provide an opportunity to further test this hypothesis.

In conclusion, our study pairs bioinformatics and cytology to validate observations of decreasing centromere repeat content through maize domestication.  We show that patterns of CentC similarity are consistent with tandem duplications and maize’s allopolyploid history has left no significant differentiation between CentC’s from the different maize subgenomes. We also identify the chromosome 4 specific repeat, cent4, as the TE RLX\_sela and show that it has increased through domestication, unlike a majority of the other TEs \citep{Chia2012}.

>>>>

\jri{check tense throughout. i think some is past and some is present} 
Our analysis of centromere repeat diversity across the maize genome identifies thousands of copies exhibiting tremendous diversity. 
Nonetheless, while we can cluster the repeats into groups of related sequences, these groups seem to have little relation to current or ancient maize chromosomes (Figure \ref{pcoa}).  
We see no evidence of chromosome specific repeats such as that seen in X and Y \citep{Kawabe2005, Pontes2004}, suggesting the presence of mechanisms homogenizing repeats across centromeres on different chromosomes. We further show that Cent4, while specific to chromosome 4, appears to be a poorly characterized tandem repeat or retroelement, but is not associated with the centromere.

Instead of chromosome-specific groups, we find that virtually all the large arrays of CentC in the maize  genome derive from the maize1 subgenome, which is known to have lower gene loss and higher average expression than the maize subgenome 2 \cite{Schnable2011}. \jri{is it maize1 or maize subgenome 1? you use both here} 
One possible explanation for this clear bias is that differential expression between subgenomes affects centromere function.  
Regulatory microRNA’s corresponding to centromeric repeats have been observed\citep{ReinhartBartel2002}, and perhaps the silenced centromere became more susceptible to deletion just like other parts of subgenome2.  
We see no evidence of a more rapid decay of CentC sequences in subgenome2, but we likely have poor ability to detect deletion of entire tandem arrays.  
An alternative explanation is may simply be that the subgenome1 parent had more or larger CentC arrays, leading to an unequal representation in the ancient polyploid ancestor of maize.

@@

%!RELATEDNESS
Our sequence comparison of CentCs also enabled us to explore the relationship between genetic and physical distance between repeats.  
Using the well-assembled centromeres on chromosomes 2 and 5, we found spatial autocorrelation among repeats, and we observe genome-wide that many CentC’s within an array fall into the same genetic cluster.  
These observations are consistent with the simple idea that most repeats arise due to tandem duplication, and long distance transposition of CentC, while necessary to homogenize repeats across chromosomes \citep{Shi2010}, is relatively uncommon.

One unusual result from our sequence comparison was the finding that pairs of CentC on different chromosomes may be identical or nearly so.  Our simulations suggest that this is plausible due to homoplasmy alone under realistic assumptions about mutation rate and divergence time.  We speculate that the vast majority of the CentC’s exist as a result of old duplications and that mutations have had a long time to accumulate. Roughly 80\% of the CentC repeats have their closest genetic relative on the same chromosome, as expected under a model of tandem duplication.  However, only 14\% of closest genetic pairs are found within 10KB of each other, suggesting that their tandem duplication is old and that most CentC’s have persisted within the genome for a long time.

%pairs investigating why a pair of chromosomes on two different chromosomes are each other closest genetic relative, we revealed that homoplasy in mutations is common across CentC variants.  We suggest that copies of CentC are sufficiently old within the genome that homoplasious mutations cause physically distant CentC’s to be highly genetically related.  Importantly, we also do not see PCoA group capturing full clusters across chromosomes in a way that would be consistent with retrotransposition.

 % Given the lack of a genome wide pattern, we also wanted to investigate genetic relatedness on individual chromosomes.  We hypothesized that most copies of CentC arose from local duplication rather than transposition, and therefore the genetic distance between CentC’s would be lowest across CentC’s within a large tandem array.  \pb{This part is ugly, and needs help} From the reference genome, we chose to investigate chromosomes 2 and 5, since they have been sequenced from end to end, allowing for inferences about the relationship between diversity and physical proximity \citep{Wolfgruber2009}.  Furthermore, the role of the repeat arrays on chromosomes 2 and 5 appear very different, as the largest array on chromosome 2 interacts with the kinetochore, while no array on chromosome 5 does \citep{Wolfgruber2009}.  We find that repeats on chromosomes 2 and 5 are most highly genetically related to neighboring copies (Figure \ref{heatmap}).  This relationship was recapitulated in analyses using SpaGeDi, where CentC’s within approximately 10-50KB of one another were more genetically similar (Tables \ref{Supp_SpagediChr02} and \ref{Supp_SpagediChr05}).  Though we did not investigate the relationship between CentC location and genetic similarity on the other chromosomes due to incomplete sequencing of centromeres, we observe that many CentC’s within an array fall into the same  significant grouping in our Tracy-Widom clustering analysis, suggesting that high local similarity of CentC’s is a genome-wide phenomenon.  This local relatedness suggests that most CentC is evolving through tandem duplication and not a long distance mechanism such as transposition that had been previously suggested \citep{Shi2010}.  Concerning chromosomes 2 and 5, the higher local relatedness of all CentC clusters regardless of presence within the kinetochore suggests that CentC interaction with kinetochore proteins does not have a detectable change on their rates of evolution.

\jri{insert CITE, italics, probably refs to tables/figs as appropriate}  Previous cytogenetic work  identified  differences in centromere repeat content  between domesticated maize and its wild relatives Z. mays ssp. parviglumis,), Z. mays ssp. mexicana,, and Z. luxurians (hereafter parviglumis, mexicana, and luxurians) \citep{Albert2010} but was unable to quantify these differences. Our resequencing results show that while there is little difference in the distribution of CentC in tandem arrays, the total abundance of CentC has  decreased  \jri{can we say absolute abundance? we have genome size estimates, right?} during domestication, and we verify this with FISH in two maize-teosinte hybrid individuals (Figure \ref{abundance}). This result contrasts with the apparent increase in transposable element abundance  \citep{Chia2012}, suggesting that changes in CentC are not due to causes common to all repetitive sequence. \jri{ are we sure TEs have increase in absolute number? didn't chia et al. look at RPKM?  wouldn't that increase if overall genome size decreases? if we control for genome size, does TE content show and absolute increase?} The maize genome is smaller than its teosinte counterpart, largely due to differences in the abundance of heterochromatic knobs (Poggio et al 1998).  \citet{Zhang2012} have postulated an adaptive relationship between centromere size and genome size based on an observed correlation between centromere size and genome size across a number of grass species.  This model would propose that the decrease in CentC abundance seen post-domestication is due to selection for smaller active centromeres to complement the smaller overall genome size.  While our current data are insufficient to evaluate this conclusion, future work investigating differences in CentC content among maize landraces that vary in genome size (CITE) may provide an opportunity to further test this hypothesis.

In conclusion, our detailed study of centromere repeats in the maize genome has highlighted differential contribution of subgenome, spatial autocorrelation along a chromosomes, and changes in abundance over the short time scale of domestication.  Further work evaluating CentC evolution across multiple populations and multiple related species may shed additional light on the timing and causes of these changes.
